---
title: "Data Storytelling - Final Project"
author: "Joey Clancy"
date: "2/23/2022"
output:
  html_document:
    code_download: yes
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r clear-environment, echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

# Data Preprocessing

```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
library(plotly)
library(kableExtra)
library(gridExtra)
library(rpart)
library(rpart.plot)
library(gghighlight)
library(knitr)
```

```{r}
# Load Data
weather <- read.csv("enchantWeather.csv")
lotto <- read.csv("enchantLotto.csv")

# Rename Day.of.Year column on lottery results set
lotto <- rename(lotto, Day = Day.of.Year)

# Merge data frames on Day column
df = merge(x = lotto, y = weather, by = "Day")

df$Preferred.Entry.Date <- as.Date(df$Preferred.Entry.Date, "%m/%d/%Y") #Convert to date type
df$Result <- as.factor(df$Result) #Convert result to factor
df$Preference.Order <- as.factor(df$Preference.Order) #Convert Preference.Order to factor
df$Preferred.Division <- as.factor(df$Preferred.Division) #Convert Preferred.Division to factor
df$ï..Table.Names <- NULL #Drop table name column

df$resultBin <- ifelse(df$Result == "Accepted", 1, 0) # Create dummy variable for Result
df$nwTrekConditions <- as.integer(ifelse(df$avgPrecipIn < 0.03 & df$avgTempMean > 52, 1, 0)) # Variable to indicate if both weather conditions are met
df$Week <- strftime(df$Preferred.Entry.Date, format = "%V") # Create Week Variable
df$Week <- as.integer(df$Week)  #Convert Week variable to integer instead of character
```

```{r}
# Data summary
summary(df)
```

```{r}
# Data structure
str(df)
```

```{r}
# First 5 entries
head(df)
```

# Univariate Graphical

## UDFs

```{r}
barBox <- function(vbl) {
  grid.arrange(
    ggplot(data = df, mapping = aes(x = {{vbl}})) +
      geom_bar() +
      theme_minimal(),
    
    ggplot(data = df, mapping = aes(x = 0)) +
      geom_boxplot(mapping = aes(y = {{vbl}})) +
      coord_flip() +
      theme_minimal()
  )
}
```

```{r}
densityBox <- function(vbl) {
  grid.arrange(
    ggplot(data = df, mapping = aes(x = {{vbl}})) +
      geom_density() +
      theme_minimal(),
    
    ggplot(data = df, mapping = aes(x = 0)) +
      geom_boxplot(mapping = aes(y = {{vbl}})) +
      coord_flip() +
      theme_minimal()
  )
}
```

## Minimum Acceptable Group Size

```{r}
barBox(Minimum.Acceptable.Group.Size)
```

## Maximum Requested Group Size

```{r}
barBox(Maximum.Requested.Group.Size)

```

## Day

```{r}
densityBox(Day)
```

## Week

```{r}
densityBox(Week)
```


## Average Precipitation (inches)

```{r}
densityBox(avgPrecipIn)
```

## Average Max Temperature (F)

```{r}
densityBox(avgTempMax)
```

## Average Mean Temperature (F)

```{r}
densityBox(avgTempMean)
```

## Average Min Temperature (F)

```{r}
densityBox(avgTempMin)
```

## Result

```{r}
df %>% 
  ggplot(aes(x = Result)) +
  geom_bar() +
  theme_minimal()
```

# Multivariate Graphical

## Average Precipitation

```{r}
precipDate <- df %>% 
  filter(Preferred.Division == "Core Enchantment Zone") %>% 
  group_by(Day) %>% 
  summarise(averagePrecipitation = mean(avgPrecipIn)) %>% 
  ggplot(aes(x = Day, y = averagePrecipitation, color = averagePrecipitation < 0.03)) +
  geom_point() +
  geom_vline(xintercept = c(182, 248), linetype='dashed') +
  geom_hline(yintercept = 0.03, linetype="dashed") +
  theme_classic() +
  labs(title = "Average Precipitation Over Time",
       subtitle = "Average precipitation bottoms out in the summer months",
       x = "Day",
       y = "Average Precipitation",
       color = "Precipitation < 0.03 in")
precipDate
ggplotly(precipDate)
```

## Average Precipitation by Zone

```{r}
facetedPrecip <- df %>% 
  group_by(Day, Preferred.Division, Result) %>% 
  summarise(averagePrecipitation = mean(avgPrecipIn)) %>% 
  ggplot(aes(x = Day, y = averagePrecipitation, color = Result)) +
  geom_point(alpha=0.45, position = "jitter") +
  facet_wrap(~ Preferred.Division) +
  labs(title = "Average Precipitation by Zone",
       subtitle = "Precipitation follows a similar pattern for all zones",
       x = "Day of Year",
       y = "Average Precipitation (inches)") +
  theme_minimal()
facetedPrecip
ggplotly(facetedPrecip)
```

## Average Temperature

```{r}
tempDate <- df %>% 
  filter(Preferred.Division == "Core Enchantment Zone") %>% 
  group_by(Day) %>% 
  summarise(averageTemperature = mean(avgTempMean)) %>% 
  ggplot(aes(x = Day, y = averageTemperature, color = averageTemperature > 52)) +
  geom_point() +
  geom_hline(yintercept = 52, linetype="dashed") +
  theme_classic() +
  labs(title = "Average Temperature Over Time (Core Zone)",
       subtitle = "Average temperature tops out in the summer months",
       x = "Day",
       y = "Average Temperature",
       color = "Temperature > 52F")

tempDate
ggplotly(tempDate)
```

## Average Temperature by Zone

```{r}
facetedTemp <- df %>% 
  group_by(Day, Preferred.Division, Result) %>% 
  summarise(averageTemp = mean(avgTempMean)) %>% 
  ggplot(aes(x = Day, y = averageTemp, color = Result)) +
  geom_point(alpha=0.45, position = "jitter") +
  facet_wrap(~ Preferred.Division) +
  labs(title = "Average Temperature by Zone",
       subtitle = "Temperature follows a similar pattern for all zones",
       x = "Day of Year",
       y = "Average Temperature (F)") +
  theme_minimal()
facetedTemp
ggplotly(facetedTemp)
```

## Acceptance Rate

```{r}
# Create separate data frame containing ratios of accepted to unsuccessful
resultRatio <- df %>% 
 group_by(Preferred.Division, Result) %>%
 summarize(N = n()) %>%
 mutate(Ratio = round(N / sum(N), 2))
```

```{r}
# Returns flipped column chart with top value highlighted
avgAcceptance <- resultRatio %>%
  filter(Result == "Accepted") %>% 
  group_by(Preferred.Division) %>% 
  summarise(avgAcceptance = mean(Ratio)) %>%  # Doesn't really do anything, as there's only one ratio for each group
  ggplot(mapping = aes(fct_reorder(Preferred.Division, avgAcceptance), y = avgAcceptance*100)) +
    geom_col() +
  coord_flip() +
  gghighlight(Preferred.Division == "Core Enchantment Zone") +
  labs(title = "Acceptance Rate by Division",
       subtitle = "Core Enchantment Zone had the lowest acceptance rate (about 1%)",
       x = "Preferred Division",
       y = "Average Acceptance (%)") +
  theme_minimal()
avgAcceptance
ggplotly(avgAcceptance)
```


## NW Trek Conditions

```{r}
df %>% 
  ggplot(aes(x = Day, y = nwTrekConditions)) +
  geom_point() +
  theme_minimal()
```

# Statistical EDA

## Train Test Split
 
```{r}
dfSpec <- df[,c('Result', 'Day', 'Preferred.Division', 'Maximum.Requested.Group.Size')]

set.seed(777)

#Use 70% of dataset as training set and remaining 30% as testing set
sample <- sample(c(TRUE, FALSE), nrow(dfSpec), replace=TRUE, prob=c(0.7,0.3))
train <- dfSpec[sample, ]
test <- dfSpec[!sample, ]  
```

## Dealing with class imbalances

```{r}
# Retrieve counts of accepted and unsuccessful applications
table(train$Result)
```


```{r}
# Over sampling
library(ROSE)
df_balanced_over <- ovun.sample(Result ~ Maximum.Requested.Group.Size + Day + Preferred.Division, data = train, method = "over", N=148726)$data
table(df_balanced_over$Result)
```

```{r}
# Under sampling
df_balanced_under <- ovun.sample(Result ~ Maximum.Requested.Group.Size + Day + Preferred.Division, data = train, method = "under", N=3444)$data
table(df_balanced_under$Result)
```

```{r}
# Using both over and under sampling
df_balanced_both <- ovun.sample(Result ~ Maximum.Requested.Group.Size + Day + Preferred.Division, data = train, method = "both", p=0.5, N=76085, seed = 1)$data
table(df_balanced_both$Result)
```

```{r}
# Synthetic data generation
df.rose <- ROSE(Result ~ Maximum.Requested.Group.Size + Day + Preferred.Division, data = train, seed=1)$data

table(df.rose$Result)
```

## Decision Trees

```{r}
tree.rose <- rpart(Result ~ ., 
                   data = df.rose, 
                   method = "class")

tree.over <- rpart(Result ~ ., 
                   data = df_balanced_over, 
                   method = "class")


tree.under <- rpart(Result ~ ., 
                    data = df_balanced_under, 
                    method = "class")

tree.both <- rpart(Result ~ ., 
                   data = df_balanced_both, 
                   method = "class")

```

### Rose Data CM

```{r}
pred.tree.rose <- predict(tree.rose, newdata=test, type = "class")
caret::confusionMatrix(pred.tree.rose, test$Result)
#plotcp(tree.rose)
```

### Oversampled Data CM

```{r}
pred.tree.over <- predict(tree.over, newdata=test, type = "class")
caret::confusionMatrix(pred.tree.over, test$Result)
#plotcp(tree.over)
```

### Undersampled Data CM

```{r}
pred.tree.under <- predict(tree.under, newdata=test, type = "class")
caret::confusionMatrix(pred.tree.under, test$Result)
#plotcp(tree.under)
```

### Over & Under Data CM

```{r}
pred.tree.both <- predict(tree.both, newdata=test, type = "class")
caret::confusionMatrix(pred.tree.both, test$Result)
#plotcp(tree.both)
```

## Final Tree - Under & Over Sampling

```{r}
rpart.plot(tree.under)
```

**Notes on Acceptance Rates**

- Before June 17th, probability of acceptance is **80%**

- After October 5th, probability of acceptance rises to **85%**

- Between October 5th and June 17th, acceptance rate is **27%**

**Notes on Preferred Division and Group Size**

- We find that maximum requested group size did not play an important role in estimating acceptance rates

- Preferred division, however, is important. If you are wanting to be selected for either the Core Zone or Colchuck zone, your acceptance probability is *heavily* dependent on time of year.


## Decision Tree - Optimal Time of Year (All Zones)

**Note about trees:**

Each node shows

- the predicted class (weather conditions met or not),

- the predicted probability of weather conditions being met,

- the percentage of observations in the node

```{r}
set.seed(777)

#Use 70% of dataset as training set and remaining 30% as testing set
sample <- sample(c(TRUE, FALSE), nrow(df), replace=TRUE, prob=c(0.7,0.3))
train <- df[sample, ]
test <- df[!sample, ]  
```


```{r}
tree <- rpart(nwTrekConditions ~ Day, train, method = "class")
rpart.plot(tree)
```

```{r}
# Get test set predictions
tree.pred <- predict(tree, test, type = "class")

# Build confusion matrix with caret package
caret::confusionMatrix(as.factor(tree.pred), as.factor(test$nwTrekConditions))
```

```{r}
# Retrieve complexity parameter or CP
printcp(tree)
```

```{r}
plotcp(tree)
```

**Notes**

- The plotcp() function provides the cross validated error rate for various complexity parameter thresholds.

- Ideally, you would expect the error to be very high for high values of cp, which will then gradually decrease before increasing again or flattening out (bias-variance trade-off).

- Observed that the complexity parameter can remain at the defaul of 0.01 as adjusting it to 0.011 produces no noticeable effect on the model or the predictions

## Decision Tree - Optimal Time of Year (Core Enchantment Zone)

```{r}
coreZoneDat <- df %>% 
  filter(Preferred.Division == "Core Enchantment Zone")
```


```{r}
set.seed(777)

#Use 70% of dataset as training set and remaining 30% as testing set
sampleCore <- sample(c(TRUE, FALSE), nrow(coreZoneDat), replace=TRUE, prob=c(0.7,0.3))
trainCore <- coreZoneDat[sampleCore, ]
testCore <- coreZoneDat[!sampleCore, ]  
```

```{r}
treeCore <- rpart(nwTrekConditions ~ Day, trainCore, method = "class")
rpart.plot(treeCore)
```

```{r}
# Get test set predictions
treeCore.pred <- predict(treeCore, testCore, type = "class")

# Build confusion matrix with caret package
caret::confusionMatrix(as.factor(treeCore.pred), as.factor(testCore$nwTrekConditions))
```

```{r}
# Retrieve complexity parameter or CP
printcp(treeCore)
```

```{r}
plotcp(treeCore)
```

# Random Forests

## RF - Original data

```{r}
library(randomForest)
# fit random forest model
rf_orig <- randomForest(
  formula = Result ~ .,
  data = train
)

rf_orig
```

```{r}
pred.forest.orig <- predict(rf_orig, newdata = test)
caret::confusionMatrix(pred.forest.orig, test$Result)
```

## On Synthetically Generated Data

```{r}
library(randomForest)
```

```{r}
set.seed(1)

# fit random forest model
rf_rose <- randomForest(
  formula = Result ~ .,
  data = df.rose
)

rf_rose
```

```{r}
which.min(rf_rose$err.rate)
```

```{r}
# Find RMSE

sqrt(rf_rose$err.rate[which.min(rf_rose$err.rate)]) 
```

```{r}
plot(rf_rose)
```

```{r}
varImpPlot(rf_rose)
```

**Notes**

- The x-axis displays the average increase in node purity of the regression trees based on splitting on the various predictors displayed on the y-axis

```{r}
rf_rose_tuned <- tuneRF(
               x=df.rose[,c(2,3,4)], #define predictor variables
               y=df.rose$Result, #define response variable
               ntreeTry=500,
               mtryStart=3, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
```

```{r}
pred.forest.rose <- predict(rf_rose, newdata = test)
caret::confusionMatrix(pred.forest.rose, test$Result)
```

## Over and Undersampled Data

```{r}
# fit random forest model
rf_both <- randomForest(
  formula = Result ~ .,
  data = df_balanced_both
)

rf_both
```

```{r}
which.min(rf_both$err.rate)
```

```{r}
# Find RMSE

sqrt(rf_both$err.rate[which.min(rf_both$err.rate)]) 
```

```{r}
plot(rf_both)
```

```{r}
varImpPlot(rf_both)
```

```{r}
rf_both_tuned <- tuneRF(
               x=df_balanced_both[,c(2,3,4)], #define predictor variables
               y=df_balanced_both$Result, #define response variable
               ntreeTry=500,
               mtryStart=3, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
```

```{r}
pred.forest.both <- predict(rf_both, newdata = test)
caret::confusionMatrix(pred.forest.both, test$Result)
```

## Undersampled Data

```{r}
# fit random forest model
rf_under <- randomForest(
  formula = Result ~ .,
  data = df_balanced_under
)

rf_under
```

```{r}
which.min(rf_under$err.rate)
```

```{r}
# Find RMSE

sqrt(rf_under$err.rate[which.min(rf_under$err.rate)]) 
```

```{r}
plot(rf_under)
```

```{r}
varImpPlot(rf_under)
```

```{r}
rf_under_tuned <- tuneRF(
               x=df_balanced_under[,c(2,3,4)], #define predictor variables
               y=df_balanced_under$Result, #define response variable
               ntreeTry=500,
               mtryStart=3, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
```

```{r}
pred.forest.under <- predict(rf_under, newdata = test)
caret::confusionMatrix(pred.forest.under, test$Result)
```

## Oversampled Data

```{r}
# fit random forest model
rf_over <- randomForest(
  formula = Result ~ .,
  data = df_balanced_over
)

rf_over
```

```{r}
which.min(rf_over$err.rate)
```

```{r}
# Find RMSE

sqrt(rf_over$err.rate[which.min(rf_over$err.rate)]) 
```

```{r}
varImpPlot(rf_over)
```

```{r}
plot(rf_over)
```

```{r}
rf_over_tuned <- tuneRF(
               x=df_balanced_over[,c(2,3,4)], #define predictor variables
               y=df_balanced_over$Result, #define response variable
               ntreeTry=500,
               mtryStart=3, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )
```

```{r}
pred.forest.over <- predict(rf_over, newdata = test)
caret::confusionMatrix(pred.forest.over, test$Result)
```

# New Data Input - Test

```{r}
#define new observation
new <- data.frame(Day=178, Preferred.Division="Snow Zone", Maximum.Requested.Group.Size=5)

#use fitted bagged model to predict Ozone value of new observation
predict(tree.under, newdata=new)
```

